---
title: "Getting and Cleaning Data Course Project"
author: "Edward May"
date: "April 19, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("data.table", lib.loc="~/R/win-library/3.4")
library(reshape2)
```

## Getting and Cleaning Data Course Project

This is a project to tidy the data from the 

Human Activity Recognition Using Smartphones Dataset
Version 1.0

Attribution and License:

Use of this dataset in publications must be acknowledged by referencing the following publication [1] 

[1] Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support Vector Machine. International Workshop of Ambient Assisted Living (IWAAL 2012). Vitoria-Gasteiz, Spain. Dec 2012

This dataset is distributed AS-IS and no responsibility implied or explicit can be addressed to the authors or their institutions for its use or misuse. Any commercial use is prohibited.

Jorge L. Reyes-Ortiz, Alessandro Ghio, Luca Oneto, Davide Anguita. November 2012.

#Loading Data
##Unzip data set
We first unzip the files creating a directory structure:

```{r}
setwd("~/Coursera/Getting and Cleaning Data/Project/UCI HAR Dataset")
dir(recursive=TRUE)
```

##Load data into variables

```{r}
x_test <- read.table("test/X_test.txt")   ## single or multiple spaces between entries are handles by read.table
features <- readLines("features.txt")
xnames <- colnames(x_test)
setnames(x_test,xnames,features)            ##sets the column names for x_test
y_test <- readLines("test/y_test.txt")
x_test$label <- y_test                 ##adds the column with test labels
subject_test <- readLines("test/subject_test.txt")
x_test$subject <- subject_test
##now change the labels to descriptive test  (there's probably an easier way to do this)
activity_labels <- read.table("activity_labels.txt")
x_test$label <- gsub("1",activity_labels[1,2],x_test$label)
x_test$label <- gsub("2",activity_labels[2,2],x_test$label)
x_test$label <- gsub("3",activity_labels[3,2],x_test$label)
x_test$label <- gsub("4",activity_labels[4,2],x_test$label)
x_test$label <- gsub("5",activity_labels[5,2],x_test$label)
x_test$label <- gsub("6",activity_labels[6,2],x_test$label)

```

Now reduce the file to only the mean and standard deviation columns:
```{r}
stdandmean <- (grepl("mean",features) | grepl("Mean",features) | grepl("std",features))
x1_test <- x_test[,stdandmean]
#View(x1_test)
```

Now repeat for the Training dataset

```{r}
x_train <- read.table("train/X_train.txt" )   ## single or multiple spaces between entries are handles by read.table
#features <- readLines("features.txt")
xnames <- colnames(x_train)
setnames(x_train,xnames,features)            ##sets the column names for x_train
y_train <- readLines("train/y_train.txt")
x_train$label <- y_train                 ##adds the column with train labels
subject_train <- readLines("train/subject_train.txt")
x_train$subject <- subject_train
##now change the labels to descriptive words  (there's probably an easier way to do this)
activity_labels <- read.table("activity_labels.txt")
x_train$label <- gsub("1",activity_labels[1,2],x_train$label)
x_train$label <- gsub("2",activity_labels[2,2],x_train$label)
x_train$label <- gsub("3",activity_labels[3,2],x_train$label)
x_train$label <- gsub("4",activity_labels[4,2],x_train$label)
x_train$label <- gsub("5",activity_labels[5,2],x_train$label)
x_train$label <- gsub("6",activity_labels[6,2],x_train$label)

```

Now reduce the file to only the mean and standard deviation columns:
```{r}
stdandmean <- (grepl("mean",features) | grepl("Mean",features) | grepl("std",features))
x1_train <- x_train[,stdandmean]
#View(x1_train)
```

now add a column indicating the data source and row bind the two data sets

```{r}
x1_test$dataset <- "test"
x1_train$dataset <- "train"
all <- rbind(x1_test,x1_train)
head(all[,1:3])
tail(all[,86:89])
```
We can look at the combinations of subjects and positions with:

```{r}
table(all$subject,all$label)
```

We need a logical list of the variable names:

```{r}
rf <- features[stdandmean]
```

then we melt and cast the data:
```{r}
submelt <- melt(all,id="subject",measure.vars = rf)
submean <- dcast(submelt,subject ~ rf,mean)
submean <- submean[order(as.numeric(submean$subject)),]         ##order them
#View(submean)

labelmelt <- melt(all,id="label",measure.vars = rf)
labelmean <- dcast(labelmelt,label ~ rf,mean)
#View(labelmean)
```

Need to change the first column to "test_data", and put in "Subject 1 averages->", or "Walking averages->"
```{r}
setnames(labelmean,"label","test data")
setnames(submean,"subject","test data")
labelmean$`test data` <- paste(labelmean$`test data`,"averages->")
submean$`test data` <- paste("Subject",submean$`test data`,"averages->")
final <- rbind(submean,labelmean)
#View(final)
head(final[,1:3])
tail(final[,1:3])
```